Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2020-02-10 rnni_convexity.tex
David's comment re Prop 1; the 'refines' definition; Figure 4 explanation

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$, $[C_1, \ldots, C_{n-2}] := R$
\FOR {$k = 1, \dots, n-2$}
    \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
        \IF {node $u$ with $\rank_{T'}(u) = rank_{T'}(\mrca_{T'}(C_k))-1$ is adjacent to $\mrca_{T'}(C_k)$}
            \STATE $T'$ is $T'$ with $\rank_{T'}(\mrca_{T'}(C_k))$ decreased by an $\nni$ move
        \ELSE
            \STATE $T''$ is $T'$ with ranks of $u$ and $\mrca_{T'}(C_k)$ swapped
        \ENDIF
        \label{alg:findpath:line:move_set_down}
        \STATE $T' = T''$
        \STATE $p = p+T'$
      \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\proof
It suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.

@@

The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n, \leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$, which means...
For simplification we will denote the partition lattice on $n$ elements by $\Pi_n$.
$\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.
This algebraic structure is related to the $\rnni$ graph in the following way.

\begin{theorem}
The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
The corresponding metric spaces are isometric.
\end{theorem}

@@

An $\rnni$ move on a tree $T$ corresponds to changing exactly one partition on $\mathcal{C}_T$ as only one cluster changes by an $\rnni$ move on $T$.
From this we can follow that the $\rnni$ graph and the metric space on maximum chains of the partition lattice with this operation are isometric.
\end{proof}

@@

>>>>

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$, $[C_1, \ldots, C_{n-2}] := R$
\FOR {$k = 1, \dots, n-2$}
\label{alg:findpath:line:for_loop}
    \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
        \IF {node $u$ with $\rank_{T'}(u) = rank_{T'}(\mrca_{T'}(C_k))-1$ is adjacent to $\mrca_{T'}(C_k)$}
            \STATE $T'$ is $T'$ with $\rank_{T'}(\mrca_{T'}(C_k))$ decreased by an $\nni$ move
        \ELSE
            \STATE $T''$ is $T'$ with ranks of $u$ and $\mrca_{T'}(C_k)$ swapped
        \ENDIF
        \label{alg:findpath:line:move_set_down}
        \STATE $T' = T''$
        \STATE $p = p+T'$
      \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\proof
It is not hard to see that the algorithm terminates with a path ending in $R$.
Indeed, every iteration of the \textbf{for} loop (line~\ref{alg:findpath:line:for_loop}) is completed once $\rank_{T'}(\mrca_{T'}(C_k)) = k$, that is, the first $k$ clusters of $T'$ and $R$ coincide.
Hence after $n-2$ such iteration the path has to arrive at $R$.

To complete the proof, it suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.

@@

The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n, \leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $X \leq Y$ if partition $X$ refines $Y$, that is, $X \leq Y \Leftrightarrow (\forall x \in X)(\exists y \in Y) x \subseteq y$.
For simplification we will denote the partition lattice on $n$ elements by $\Pi_n$.
$\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
We assume that a partition $X$ in the partition lattice $\Pi_n$ has rank $k$ if the number of elements in $X$ is $n-k$.
The algebraic structure of $\Pi_n$ is related to the $\rnni$ graph on $n$-taxa trees in the following way.

\begin{theorem}
The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
The corresponding metric spaces are isometric.
\label{thm:partition_lattice}
\end{theorem}

@@

Note that this bijection is an isomorphism between the $\rnni$ graph and the graph of chains as in the theorem.
Indeed, two chains are different by exactly one partition if and only if the corresponding trees are connected by an $\rnni$ move.
\end{proof}

@@

Figure~\ref{fig:partition_lattice4} is an illustration of the proof of Theorem~\ref{thm:partition_lattice}.
The four chains indicated in bold correspond to the following $\rnni$ path.
The leftmost chain corresponds to the caterpillar tree $[1, 2, 3, 4]$.
First, the partition $\{1, 2, 3\} \{4\}$ is replaced with $\{1, 2\} \{3, 4\}$ and we get the chain corresponding to the tree $[\{1, 2\}, \{3, 4\}]$ (in the cluster representation), which is one $\rnni$ move away from the caterpillar tree.
Second, the partition $\{1, 2\} \{3\} \{4\}$ is replaced with $\{1\} \{2\} \{3, 4\}$, which corresponds to the rank swap on the previous tree.
Third, the partition $\{1, 2\} \{3, 4\}$ is replaced with $\{1\} \{2, 3, 4\}$ and we reach the caterpillar tree $[3, 4, 2, 1]$.
